Analysing Constraint Grammar with a SAT solver

### 2015-10-19 paper/lrec/implementation.tex
text

\begin{table*}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Word} & \textbf{Variables}                                                          & \textbf{Default rules}               & \textbf{Constraint rules}                                  \\ \hline
``the''         & \texttt{the\_det}                                                           & \texttt{the\_det}                    & \multirow{2}{*}{\texttt{the\_det}  $\Rightarrow$ $\neg$\texttt{bear\_v\_pl}} \\ \cline{1-3}
``bear''        & \begin{tabular}[c]{@{}l@{}}\texttt{bear\_n\_sg} \\ \texttt{bear\_v\_pl}\end{tabular} & \texttt{bear\_n\_sg} $\vee$ \texttt{bear\_v\_pl} &                                                            \\ \hline
\end{tabular}
\caption{SAT-encoding of \texttt{REMOVE v IF (-1 det)} applied to \emph{the bear}}
\label{table:vars}
\end{table*}

@@

In this section, we describe the implementation of the toolkit.
We encode CG rules as a SAT problem, as in \cite{listenmaa_claessen2015}, and from the morphological lexicon, we get constraints for individual words. This allows us to create word sequences that will or will not trigger a given rule or rule set.

@@

@@

Given a large morphological lexicon, it is relatively easy to restrict each individual word to contain only realistic ambiguities. We expand the lexicon, map each wordform to its possible analyses, and then ignore the actual wordforms. For instance, a Spanish morphological lexicon contains the entries \texttt{casa:<verb><sg><p3>} and \texttt{casa:<noun><sg>}, hence we know that a confusion between a 3rd person singular verb and a singular noun is possible. The lexicon is not likely to contain a wordform that can be analysed both as an adjective and punctuation, so that kind of ambiguous word is never created.

>>>>

@@

In this section, we describe the implementation of the toolset.
We encode CG rules as a SAT problem, as in \cite{listenmaa_claessen2015}, and from the morphological lexicon, we get constraints for individual words. This allows us to create word sequences that will or will not trigger a given rule or rule set.

@@

% table:vars moved to main file for positioning

@@

Given a large morphological lexicon, we can restrict each individual word to contain only realistic ambiguities. We expand the lexicon, map each wordform to its possible analyses, and then ignore the actual wordforms. For instance, a Spanish morphological lexicon contains the entries \texttt{casa:<verb><sg><p3>} and \texttt{casa:<noun><sg>}, hence we know that a confusion between a 3rd person singular verb and a singular noun is possible. 
%The lexicon is not likely to contain a wordform that can be analysed both as an adjective and punctuation, so that kind of ambiguous word is never created.

### 2015-10-19 paper/lrec/lrec2016_listenmaa_claessen.tex
text

\documentclass[11pt]{article}
\usepackage{nodalida2015}
\usepackage{times}
\usepackage{mathptmx}
\usepackage{fixltx2e}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{multirow}
\usepackage{latexsym}
\usepackage{cite}
%\usepackage{graphics}
\usepackage{graphicx}
\usepackage{float}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{authordate1-4}
\usepackage{multirow}
%\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\begin{abstract}
We describe a method for analysing Constraint Grammar.
We encode a grammar as a SAT problem and find out
whether there are conflicting or redundant rules.
With a morphological lexicon, we can construct example sequences that
have realistic ambiguities.
We test our tools with grammars from 100-1000 rules.
\end{abstract}

>>>>

\documentclass[11pt]{article}
\usepackage{nodalida2015}
\usepackage{times}
\usepackage{mathptmx}
\usepackage{fixltx2e}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{multirow}
\usepackage{latexsym}
\usepackage{cite}
%\usepackage{graphics}
\usepackage{graphicx}
%\usepackage{float}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{authordate1-4}
\usepackage{multirow}
%\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\begin{abstract}
We describe a method for analysing Constraint Grammar by encoding it
as a SAT problem.
Our tools can detect internal conflicts or redundancies in a grammar,
as well as generate example sequences to demonstrate the effect of some rule or rule set.
No corpus is required, only a morphological lexicon.
%We test our tools with grammars from 100-1000 rules.
\end{abstract}

### 2015-10-19 paper/lrec/problem_desc.tex
more text
text

Constraint Grammar (CG, \cite{karlsson1995constraint})
% \citep[CG][]{karlsson1995constraint} 
is a method for disambiguating morphologically analysed text% and adding syntactic or dependency labels
. 
A grammar consists of rules that target specific analyses for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1C det) ;}
\end{itemize}
removes a verb reading from a word which is preceded by an unambiguously (C) tagged determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<bear>"
        "bear" noun sg
        "bear" verb pl
\end{verbatim}
\end{itemize}
the rule will match to the word \emph{bear}, and remove the analysis \texttt{verb pl}.
However, if the target word has only one remaining analysis, then the rule will not apply, even if the condition is met.

@@

\begin{itemize}
\item Are there rules that contradict each other?
\item Are there rules that will never fire?
\item Generate a sequence that triggers rule(s) X but not rule(s) Y
\item Generate a sequence that is ambiguous but doesn't trigger any rules
\end{itemize}

@@

In addition, there is a large body of research on automatically inducing rules, e.g. \cite{inducing_cg1996}, \cite{lindberg_eineborg98ilp}%, \cite{lager01transformation}, \cite{asfrent14} 
. However, since our work is aimed to aid the process of hand-crafting rules, we omit those works from our discussion.

@@

Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin \cite{voutilainen2004}.
%\cite{voutilainen2004} gives a detailed account about best practices of grammar writing and efficient use of corpora to aid the grammar development.
%For a language with no free or tagset-compatible corpus available, \cite{tyers_reynolds2015} describe a method where they apply their rules to unannotated Wikipedia texts and pick 100 examples at random for manual check.

CG rules are usually arranged in sections, and run in the following manner. 
First apply rules from section 1, and repeat until nothing changes in the text. Then apply rules from sections 1--2, then 1--3 and so on, until the set includes all rules.
The best strategy is to place the safest and most effective rules in the first sections,
so that they make way for the following, more heuristic and less safe rules to act on.

A representative corpus is arguably the best way to get concrete numbers---how many times a rule applied and how often it was correct---and to arrange the rules in sections based on that feedback.
However, this method will not notice a missed opportunity or a grammar-internal conflict, nor suggest ways to improve.
%, for instance changing the place of a rule.

% \cite{voutilainen2004} state that the around 200 rules are probably enough to resolve 50--75 \% of ambiguities in the corpus used in the development. 
% This is invaluable, especially at the beginning stages of grammar writing. 
% Sometimes you just want to know that a rule fired 500 times and was correct 490 times, rather than ``it is possible to construct a word sequence that will trigger this rule''.

%The connection to the reality has also its limitations. 
% What if instead of firing 500 times, it could have fired 1000 times if it had been placed earlier in the rule sequence.
%Or the rule that was wrong in 60 \% of the time,

\paragraph{Optimising hand-written grammars}
The corpus-based method can tell the effect of each single rule at their place in the rule sequence, and it is up to the grammar writer to make changes in the grammar.
\cite{bick2013tuning} automates the latter process: the rules are moved into different sections or removed in total based on their effect, and in parallel, their contexts are made stricter or less strict.
After trying out several combinations, the version that produces the best results is kept.

This is a valuable tool for handling grammars that are too big to keep track of, and Bick's evaluation shows that the tuned grammars perform better than the original.
As a downside, the grammar writer will likely not know why exactly is the tuned grammar performing better.
% At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
% This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.

>>>>

Constraint Grammar (CG, \cite{karlsson1995constraint})
% \citep[CG][]{karlsson1995constraint} 
is a method for disambiguating morphologically analysed text
% and adding syntactic or dependency labels
. 
A grammar consists of rules that target specific analyses for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1C det) ;}
\end{itemize}
removes a verb reading from a word which is preceded by an unambiguously (C) tagged determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<bear>"
        "bear" noun sg
        "bear" verb pl
\end{verbatim}
\end{itemize}
the rule will match to the word \emph{bear}, and remove the analysis \texttt{verb pl}.
However, if the target word has only one remaining analysis, then the rule will not apply, even if the condition is met.

@@

\begin{itemize}
\item Are there rules that contradict each other?
\item Are there rules that will never fire?
\item Generate a sequence that triggers rule(s) X but not rule(s) Y
%\item Generate a sequence that is ambiguous but doesn't trigger any rules
\end{itemize}

@@

In addition, there is a large body of research on automatically inducing rules, e.g. \cite{inducing_cg1996}, \cite{lindberg_eineborg98ilp}
%, \cite{lager01transformation}, \cite{asfrent14} 
. However, since our work is aimed to aid the process of hand-crafting rules, we omit those works from our discussion.

@@

Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin \cite{voutilainen2004}.
This helps the grammar writer to arrange the rules in appropriate sections, from safest and most effective to less certain and unusual. 
However, this method will not notice a missed opportunity or a grammar-internal conflict, nor suggest ways to improve.

% -- uncomment for full version ?

%\cite{voutilainen2004} gives a detailed account about best practices of grammar writing and efficient use of corpora to aid the grammar development.
%For a language with no free or tagset-compatible corpus available, \cite{tyers_reynolds2015} describe a method where they apply their rules to unannotated Wikipedia texts and pick 100 examples at random for manual check.

% CG rules are usually arranged in sections, and run in the following manner. 
% First apply rules from section 1, and repeat until nothing changes in the text. Then apply rules from sections 1--2, then 1--3 and so on, until the set includes all rules.
% The best strategy is to place the safest and most effective rules in the first sections,
% so that they make way for the following, more heuristic and less safe rules to act on.

% A representative corpus is arguably the best way to get concrete numbers---how many times a rule applied and how often it was correct---and to arrange the rules in sections based on that feedback.
% However, this method will not notice a missed opportunity or a grammar-internal conflict, nor suggest ways to improve.

% \cite{voutilainen2004} state that the around 200 rules are probably enough to resolve 50--75 \% of ambiguities in the corpus used in the development.

\paragraph{Automatic optimisation of hand-written grammars }
% The corpus-based method can tell the effect of each single rule at their place in the rule sequence, and leaves the grammar writer to make changes in the grammar.

\cite{bick2013tuning} modifies the grammar automatically, by trying out different orderings of the rules, and changing their contexts to stricter or less strict.
Bick reports error reductions of 7--15\% compared to the original grammars.
However, the grammar writer will likely not know why exactly is the tuned grammar performing better.
% At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
% This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.
